---
title: "IBD Analysis - Reviewer Questions"
author: "Kevin Williams"
lab:    "Barbados Nutrition Study"
date:   "January 31, 2021"
output: 
  html_document:
    theme: default



---

```{r SETUP-KNITR, setup, include=FALSE}
      knitr::opts_chunk$set(echo = TRUE)

```
  
<br>

## Introduction

This is a brief workflow highlighting the follow-up analysis to address reviewer concerns related to the pending manuscript, ***"Assessment of the inclusion of female racial/ethnic minority, female, and older individuals in inflammatory bowel disease clinical trials"*** **(Rabinowitz, Silver et al.)**. Where appropriate, samples of the exact R syntax used will be displayed, along with the corresponding output (tabular data, graphical plots, maps, etc.).

<br>

***

### Main Questions

* What is the distribution of IBD clinical trials by "Phase?"
* For Phase 3 and Phase 4 IBD Clinical trials, how many included pregnant subjects?

<br>

***

#### Task 1:  Remediate the R data frame, "IBD" to include a factor variable called PHASES2 - a recasting of the PHASES variable as factor with five levels, including NA values

<br>


```{r FACTOR-PHASE-VARIABLE}


require(broom)
require(dplyr)

IBD3  <-
    IBD2 %>% 
    select( CATEGORY, PHASES, ANALYZED, FEMALES, PREGNANT_EXCL) %>% 
    mutate( PHASES = ifelse(is.na(PHASES), 0, PHASES)) %>% 
    mutate( PHASES2 = factor( PHASES,  levels= c(1,2,3,4,0))) %>% 
    mutate( PHASES2 = recode_factor( PHASES2, '1'= 'P1', '2'= 'P2', '3' = 'P3', '4' = 'P4', '0' = 'PNA', .ordered = T) )


```
  

<br>

Here's a glimpse of the structure of the resulting dataset `IBD3`:
  
  
```{r GLIMPSE-1 }
  
glimpse(IBD3)

```


***

#### Task 2:  Create cross-tabulations that show the distribution of studies by values of the 'PHASES2' factor variable.

<br>

```{r CROSSTAB-1 }

require(gmodels)
  
#Inlcudes Phases with an NA value (no assignment)

CrossTable(IBD3$PHASES2, IBD3$PREGNANT_EXCL, prop.chisq=F, prop.t=F)  #rows then/over columns


#Includes Phases only with a value (NA values excluded)

IBD3A <-
IBD3 %>% filter( PHASES2 != "PNA") 
  CrossTable( IBD3A$PHASES2, IBD3A$PREGNANT_EXCL, prop.chisq=F, prop.t=F)  #rows then/over columns


```


***

#### Task 3:  Create cross-tabulations based on a subset of the values of the 'PHASES2' factor variable - one that includes ONLY Phase 3 & Phase 4 trials

<br>

```{r CROSSTAB-2 }

require(gmodels)
  


IBD3B <-
IBD3 %>% filter( PHASES2 %in% c('P3', 'P4')) 
  CrossTable( IBD3B$PHASES2, IBD3B$PREGNANT_EXCL, prop.chisq=F, prop.t=F)  #rows then/over columns


```

***

#### Task 4:  Previously run cross-tabulations provide only frequency statistics for trials.  In case summary stats are requires based on the number of participants, create a series of summary tables that provide this metric.

<br>

```{r SUMMARYIES }

require(kableExtra)
  

  
IBD3_SUMMARY <- 
  
    IBD3 %>% group_by( PHASES2 ) %>% 
      summarise( COUNT = n(),
                 SUBJECTS = sum(ANALYZED)) %>% 
      mutate( PCT_N = percent(COUNT / sum(COUNT)),
              PCT_SUBJ = percent(SUBJECTS / sum(SUBJECTS))) %>%  tibble::as_tibble()

IBD3A_SUMMARY <- 
  
  IBD3A %>% group_by( PHASES2 ) %>% 
  summarise( COUNT = n(),
             SUBJECTS = sum(ANALYZED)) %>% 
  mutate( PCT_N = percent(COUNT / sum(COUNT)),
          PCT_SUBJ = percent(SUBJECTS / sum(SUBJECTS))) %>%   tibble::as_tibble()



IBD3B_SUMMARY <- 
  
  IBD3B %>% group_by( PHASES2, PREGNANT_EXCL ) %>% 
  summarise( COUNT = n(),
             SUBJECTS = sum(ANALYZED)) %>% 
  mutate( PCT_N = percent(COUNT / sum(COUNT)),
          PCT_SUBJ = percent(SUBJECTS / sum(SUBJECTS))) %>%   tibble::as_tibble()



kbl(IBD3_SUMMARY, booktabs = T, format= "html", 
    caption = "<center><strong>IBD Clinical Trials by Phase Type</strong></center>",
    align= c('l','r','r','r','r'), escape=F, 
    col.names = linebreak(c("Phase", "Trials<br>N", "Subjects<br>N", "Trials<br>Pct", "Subjects<br>Pct"))) %>% 
  
  kable_styling(full_width = F, "hover") %>% column_spec(1, width = "4cm") %>%
  scroll_box(height = "6in")



kbl(IBD3A_SUMMARY, booktabs = T, format= "html", 
    caption = "<center><strong>IBD Clinical Trials by Phase Type (NO NA)</strong></center>",
    align= c('l','r','r','r','r'), escape=F, 
    col.names = linebreak(c("Phase", "Trials<br>N", "Subjects<br>N", "Trials<br>Pct", "Subjects<br>Pct"))) %>% 
  
  kable_styling(full_width = F, "hover") %>% column_spec(1, width = "4cm") %>%
  scroll_box(height = "6in")


kbl(IBD3B_SUMMARY, booktabs = T, format= "html", 
    caption = "<center><strong>IBD Phase 3/4 Clinical Trials by Pregnancy Status</strong></center>",
    align= c('l','c','r','r','r','r'), escape=F, 
    col.names = linebreak(c("Phase", "Includes<br>Pregnant?", "Count<br>N", "Subjects<br>N", "Trials<br>Pct", "Subjects<br>Pct"))) %>% 
  
  kable_styling(full_width = F, "hover") %>% column_spec(1, width = "4cm") %>%
  scroll_box(height = "6in")


```